Discovering constraints in spreadsheets and tabular data

### 2016-09-08 Paper/ECAI/tabular.tex
Added explanation for number of vectors

For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of size~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast, the amortized time is constant in the size of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
Synthetic experiments with random integers and strings show that increasing the size of the vectors has a negligible impact on the performance for most templates.

Aggregate constraints, e.g. sum or max, have an argument that takes subblocks of varying sizes.
Therefore, the size of the blocks in the spreadsheet influences the runtime of these constraints.
For a block of size~$m$ there are $\frac{m(m + 1)}{2} = O(m^2)$ subblocks, thus, if a spreadsheet has smaller sized blocks, the search for aggregate constraints will be faster.

Synthetic experiments with random integers confirm that splitting a constant number of vectors into smaller blocks improves the performance of the aggregate constraint search.

>>>>

For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of size~$k$, the worst case complexity becomes $O(n * m * k)$.
However, in practice, most subassignments will not satisfy the constraint template.
Since our implementation tries to account for this by failing fast, the amortized time is constant in the size of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small). 
The benchmark experiments show that the number of cells in a spreadsheet has a very limited impact on the running time.
Synthetic experiments with random integers and strings show that increasing the size of the vectors has a negligible impact on the performance for most templates.

On the contrary, the number of vectors in a spreadsheet have a significant impact on the running time.
Many constraints take single vector arguments, therefore, the number of subassignments to be tested depends on the number of vectors in the spreadsheets and the arity of the constraint.
For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different).
As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a large arity.
This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors is increased.
However, in our benchmark experiments there will be various factors that limit the rapid increase in running time because the assignment step can disqualify entire blocks that do not match the signature requirements, i.e. blocks of vectors with different lengths for the product constraint.

Aggregate constraints, e.g. sum or max, have an argument that takes subblocks of varying sizes.
Therefore, the size of the blocks in the spreadsheet also influences the runtime of these constraints.
For a block of size~$m$ there are $\frac{m(m + 1)}{2}$ or $O(m^2)$ subblocks, thus, if a spreadsheet has more but smaller sized blocks, the search for aggregate constraints will be faster.
Consider, for example, $v$~vectors distributed over $b$~blocks of size $v/b$, the number of subblock candidates then is $O(b * v^2 / b^2) = O(v^2 / b)$.
Synthetic experiments with random integers confirm that splitting a constant number of vectors into smaller blocks improves the performance of the aggregate constraint search.
